---
title: "Extract Spatial Control"
author: "JMR"
toc: true
toc-expand: 1
toc-depth: 2
format:
  html:
    self-contained: true
    code-tools: true
    toc-location: right-body
  docx: 
    toc-location: body
editor: source
editor_options:
  chunk_output_type: console
vignette: >
  %\VignetteIndexEntry{Vignette's Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---


## Intro
The goal of this tutorial is to guide the user extract raster data to be use as spatial control in micro economic analysis .

## Overview of Steps
In this guide, we will go through the following steps:

1. Load the data
2. Prepare the survey coordinate
3. Extract the spatial observations based on the coordinates
4. Save the result

## Is this guide for me?

This guide provides a step-by-step approach to extract raster data based on survey locations. The target audience includes economists who may have experience with statistical software (e.g. STATA) but are less familiar with spatial data processing in R.

The document is not meant to be a course on R or on how the functions work. It is just a practice example on how to extract raster data based on coordinate location. This is done by using specific functions that wrap up as many steps as possible to ensure it is easier for the user to follow.

## What do I need before starting?

The following R packages are necessary: `terra`, `tidyverse`, `haven`, `data.table`. To install the above package you can use `install.packages("name_of_package")`, don't forget the `"`. We also need the  package `climate4economist`, which can be installed only locally.

If you are not familiar with R check the [appendix] for understanding some coding style used in this tutorial.

## Code
### Set Up
```{r paths}
#| label: paths

path_to_sp_control <- file.path(#"..",
                                "data", "spatial")
path_to_urca <- file.path(path_to_sp_control, "URCA", "URCA.tif")

path_to_survey <- file.path(#"..",
                            "data", "survey", "LSMS_labor_survey",
                            "labour_forYR_coordinates_new.dta")

path_to_sp_control <- file.path("data", "result", "lbr_srv_sp_control.dta")
```

```{r package}
#| label: package

library(climatic4economist)
```


### Read
```{r read_control}
#| label: read_control

urca <- terra::rast(path_to_urca)

```

```{r read_survey}
#| label: read_survey

surveys <- haven::read_dta(path_to_survey) |>
  dplyr::filter(survey != "Uganda16") |> # 100849
  dplyr::filter(!is.na(lon_mod) & !is.na(lat_mod)) |>
  dplyr::select(survey, hhid, id_code, lat_mod, lon_mod) |>
  dplyr::distinct(survey, hhid, .keep_all = TRUE) |>
  dplyr::group_by(survey) |>
  dplyr::group_split() |>
  setNames(c("Ethiopia19", "Malawi17", "Niger14", "Nigeria19", "Tanzania13"))
```

### Prepare the survey coordinates
```{r geo_ref}
#| label: geo_ref

surveys_id <- lapply(surveys,
                     prepare_coord,
                     lon_var = lon_mod,
                     lat_var = lat_mod) 

survey_geo <- lapply(surveys_id,
                     georef_coord,
                     geom = c("lon_mod", "lat_mod"),
                     crs = "EPSG:4326") 
```

### Crop
```{r crop}
#| label: crop

urca_cntry <- sapply(survey_geo,
                     terra::crop,
                     x = urca,
                     snap = "out",
                     simplify = FALSE)
```


### Extract
```{r extract}
#| label: extract

urca_coord <- urca_cntry |>
  purrr::map(terra::focal,
             w = 3, 
             fun = "mean", 
             na.policy = "only") |>
  purrr::map2(survey_geo,
              extract_by_coord) |>
  purrr::map(~dplyr::rename(.x, URCA = focal_mean))

```

### Merge with Survey
```{r merge_with_survey}
#| label: merge_with_survey

urca_hh <- purrr::map2(surveys_id,
                       urca_coord,
                       merge_with_survey)
```

### Write
```{r write}
#| label: write

urca_hh |>
  dplyr::bind_rows() |>
  dplyr::select(-c(lat_mod, lon_mod, x_cell, y_cell)) |>
  haven::write_dta(path_to_sp_control)
```
